 
### States pollution Index in 2017 

========================================================
      
by Saravanan Natarajan

========================================================


### Abstract
The Annual summary file from Environmental Protection Agency's Air Quality 
System for year 2017 consists of various data sets related to pollution.
Ground-level ozone and airborne particles are the two pollutants 
that pose the greatest threat to human health. In this project we will 
explore the good and sensitive days in the year 2017 for the various States
in USA.


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
suppressMessages(library(reshape))
suppressMessages(library(reshape2))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(maps))
suppressMessages(library(RColorBrewer))
suppressMessages(library(GGally))
suppressMessages(library(scales))
suppressMessages(library(memisc))
suppressMessages(library(gridExtra))
suppressMessages(library(tidyverse))
suppressMessages(library(ggthemes))
```

### Dataset
Get the data from the CSV file and explore

```{r echo=FALSE, warning=FALSE, Load_the_Data}
air_quality_data <- read.csv('annual_aqi_by_county_2017.csv')
str(air_quality_data)
```

## Introduction to dataset
The data set consists of 1044 observations with 19 variables for the year 2017.

The complete details about dataset is provided in
https://www.epa.gov/outdoor-air-quality-data/about-air-data-reports#aqi

Each row of the AQI Report lists summary values for one year for one county.

https://www.airnow.gov/index.cfm?action=aqibasics.aqi

The summary values include both qualitative measures (days of the year having 
"good" air quality, for example) and descriptive statistics (median AQI value, 
for example).

## Univariate analysis 

```{r echo=FALSE, warning=FALSE, Univariate_Summary}
summary(air_quality_data)
names(air_quality_data)
```

The summary shows various states and their counties for the year 2017, with 19
variable model fitting functions.

# AQI Days plots

```{r echo=FALSE, warning=FALSE, fig.width=18, fig.height=9, AQI_Days}
p1 <- ggplot(aes(Good.Days), data=air_quality_data) +
  geom_histogram(binwidth = 5, fill="green", color="black")

p2 <- ggplot(aes(Moderate.Days), data=air_quality_data) +
  geom_histogram(binwidth = 1, fill="orange", color="black")

p3 <- ggplot(aes(Unhealthy.for.Sensitive.Groups.Days), data=air_quality_data) +
  geom_histogram(binwidth = 0.5, fill="blue", color="black")

p4 <- ggplot(aes(Unhealthy.Days), data=air_quality_data) +
  geom_histogram(binwidth = 1, fill="yellow", color="black")

p5 <- ggplot(aes(Very.Unhealthy.Days), data=air_quality_data) +
  geom_histogram(binwidth = 0.5, fill="purple", color="black")

p6 <- ggplot(aes(Hazardous.Days), data=air_quality_data) +
  geom_histogram(binwidth = 0.5, fill="red", color="black")

grid.arrange(p1, p2, p3, p4, p5, p6, ncol= 2)
```


Histogram plot was used to view the various types of AQI, counts and range.
The binwith and color was chosen based on the number of days in the year
and AQI value of days

# Pollutant measured

A daily index value is calculated for each air pollutant measured. 
The highest of those index values is the AQI value, and the pollutant 
responsible for the highest index value is the "Main Pollutant." 
These columns give the number of days each pollutant measured was the main 
pollutant. A blank column indicates a pollutant not measured in the 
county or CBSA.

```{r echo=FALSE, warning=FALSE, fig.width=18, fig.height=9, Pollutant_measured}
p1 <- ggplot(air_quality_data, aes(Days.CO)) +
  geom_histogram(binwidth = 10, fill="yellow", color="black")

p2 <- ggplot(air_quality_data, aes(Days.NO2)) +
  geom_histogram(binwidth = 10, fill="purple", color="black")

p3 <- ggplot(air_quality_data, aes(Days.Ozone)) +
  geom_histogram(binwidth = 5, fill="red", color="black")

p4 <- ggplot(air_quality_data, aes(Days.SO2)) +
  geom_histogram(binwidth = 10, fill="orange", color="black")

p5 <-ggplot(air_quality_data, aes(Days.PM2.5)) +
  geom_histogram(binwidth = 5, fill="blue", color="black")

p6 <- ggplot(air_quality_data, aes(Days.PM10)) +
  geom_histogram(binwidth = 10, fill="green", color="black")

grid.arrange(p1, p2, p3, p4, p5, p6, ncol= 2)
```

The graph infers that Ozone and PM2.5 were the Main pollutants. CO was least 
measured pollutant in the county.

# Simple Univariate challenge

https://leanpub.com/exdata

Which counties in the United States have the highest levels of ozone pollution
in year 2017?

Get the data directly by plotting Days.with.AQI and State, choose Days.with.AQI, 
since it is the highest of daily index value calculated for each air pollutant 
measured.

```{r echo=FALSE, warning=FALSE, Days.with.AQI }
ggplot(aes(Days.with.AQI,air_quality_data$State), data=air_quality_data) +
  geom_point(alpha = 0.5, size = 1, shape=21, position = 'jitter', fill=I("#F79420")) 
```


Create variable called state_ranking with log 10 of Days.with.AQI. 

```{r echo=FALSE, warning=FALSE, Ranking}

state_ranking <- group_by(air_quality_data, State, County) %>%
  summarize(air_quality_data_log10 = log10(Days.with.AQI)) %>%
  as.data.frame %>%
  arrange(desc(air_quality_data_log10))

str(state_ranking)

```


Highly polluted States list

```{r echo=FALSE, warning=FALSE, Ranking_head}
head(state_ranking, 10)
```


Least polluted States list

```{r echo=FALSE, warning=FALSE, Ranking_tail}
tail(state_ranking, 10)
```

Interestingly California is both in the list of high and low polluted state.
The county Del Norte in California is least polluted.
Counties Fresno, Kings, and Riverside in California were highly polluted.


## Bivariate analysis

The missing parameter in the dataset is the mean value of Good and Sensitive
days in the year 2017

```{r echo=FALSE, warning=FALSE, Mean_Good_Sensitive_AQI}
air_quality_data.pc_AQI <- air_quality_data %>% 
  group_by(State, County, Days.with.AQI) %>%
  summarize(mean_Good_AQI = mean(Good.Days/Days.with.AQI),
            mean_Sensitive_AQI = mean(Unhealthy.for.Sensitive.Groups.Days/Days.with.AQI)
            )

air_quality_data.pc_AQI <- arrange(air_quality_data.pc_AQI, State, County, Days.with.AQI)

head(air_quality_data.pc_AQI, 20)
```

# Plots

The values can be plotted against the number of days in the year having an AQI
value

```{r echo=FALSE, warning=FALSE, Mean_Good_Sensitive}
p1 <- ggplot(aes(x = mean_Good_AQI, y = Days.with.AQI), data = air_quality_data.pc_AQI) +
  geom_jitter(alpha=.5, shape=21, fill=I("#F79420"))

p2 <- ggplot(aes(x = mean_Sensitive_AQI, y = Days.with.AQI), data = air_quality_data.pc_AQI) +
  geom_jitter(alpha=.5, shape=21, fill=I("#F71420"))

grid.arrange(p1, p2, ncol=1)
```


The plot provides the outlier values of both good days AQI and the sensitive 
days AQI.

## Multivariate analysis

# Sensitive group days vs Good days in each State

```{r echo=FALSE, warning=FALSE, fig.width=12, fig.height=6, pair_plots}
ggplot(air_quality_data, aes(Good.Days, Unhealthy.for.Sensitive.Groups.Days)) +
  geom_point(aes(color = State), size = 3, alpha = 1/2) 
```

Scatterplot help us to view the number of Sensitive group days vs Good days 
for each State.


# Integrate with States Map

Back to our least polluted and most polluted states, in this section we will
plot the data in US map, the main challenge is EPA data don't have the 
states lat and long, finally got the answer from

https://ekburchfield.files.wordpress.com/2017/05/web_scraping_in_r.pdf

https://s3.amazonaws.com/udacity-hosted-downloads/ud651/GeographyOfAmericanMusic.html

https://github.com/tidyverse/tidyverse/issues/66

# Map data with State abbreviation

```{r echo=FALSE, warning=FALSE, map_data_with_state_abbrevation} 
data(state)
states <- data.frame(state.abb, state.name, state.area, state.center,
                     state.division, state.region, state.x77)

states <- subset(states, ! state.abb %in% c("HI", "AK"))

head(states, 5)
```

# Map data with county details

```{r echo=FALSE, warning=FALSE, map_data_with_county}
map_usa <- maps::map("county",".", exact = FALSE, plot = FALSE, fill = TRUE) %>%
  fortify()

#Rename the region and subregion to State and County
map_usa$State <- str_to_title(map_usa$region)
map_usa$County <- str_to_title(map_usa$subregion)
head(map_usa)
```

# Merge mean_Good_AQI data with map data

```{r echo=FALSE, warning=FALSE, Merge_Good_AQI_Map_data}
states_AQI <- merge(air_quality_data.pc_AQI, map_usa, by = c("State", "County"))

states_AQI <- states_AQI[order(states_AQI$order), ]

head(states_AQI,20)
```

# Plot the data

```{r echo=FALSE, warning=FALSE, Merge_Good_AQI_Map_data_plot}

ggplot(data = states_AQI, aes(x = long, y = lat, group = group)) +
  geom_polygon() + 
  geom_path(colour = 'white', linestyle = 2) + 
  coord_map() +
  theme(legend.position = "bottom") +
  geom_text(data = states, aes(x = x, y = y, label = state.abb, group = NULL), 
            size = 4, 
            colour="red") +
  geom_polygon(aes(fill = mean_Good_AQI)) +
  labs(fill = "") +
  guides(fill = guide_colorbar(barwidth = 10, barheight = 0.5)) +
  ggtitle("Percent Good Days of total measured days")

```

Plotted the county boundaries with the state name, the region most polluted was
captured as darker region it means lower percentage of good days in the year.
The colour scales were captured in the bottom.

### Correlation between Good days AQI and Sensitive Days AQI

Complete by finding the correlation between good AQI days and 
the sensitive AQI days.


```{r echo=FALSE, warning=FALSE, correlation_value}

corr_Value = as.character(round(cor(states_AQI$mean_Good_AQI,states_AQI$mean_Sensitive_AQI), 2))

corr_Value
```

The correlation value r = -0.66


### Final Plots and Summary

Number of days in the year having an AQI value 0 through 50 means the good days
for each counties. Similar AQI days in each counties will help us to get the 
healthy states in USA. Florida and District Of Columbia were chosen to plot
the good AQI values for each counties.

http://paulorenato.com/index.php/171


```{r echo=FALSE, warning=FALSE, states_pair_plots}

air_quality_data_Florida <- filter(air_quality_data, State == "Florida")
air_quality_data_District_Of_Columbia <- filter(air_quality_data, State == "District Of Columbia")

p1 <- ggplot(aes(x = Good.Days), data = air_quality_data_Florida) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~State, scales="free") + 
  labs(x = "Number of days in the year having Good AQI", 
       y = "Similar AQI days in counties",
      title = "Good AQI values for counties in Florida ")

p2 <- ggplot(aes(x = Good.Days), data = air_quality_data_District_Of_Columbia) +
  geom_histogram(binwidth = 0.1) +
  facet_wrap(~State, scales="free") + 
  labs(x = "Number of days in the year having Good AQI", 
       y = "Similar AQI days in counties",
      title = "Good AQI values for counties in District Of Columbia")

grid.arrange(p1, p2, ncol = 1)
```

Binwidht and the scales in facet_wrap helped to get the clear 
view of good AQI days in each county.

In the above graph Florida have the maximum y scale of 4, which means 4 
counties have 131 good AQI days, the counties were Miami-Dade, Palm Beach, 
Pinellas and Sarasota. Same applicable for 149 good AQI days.

District Of Columbia state only have one county with 157 good AQI days in
the year 2017.


### Map plotting of Sensitive AQI days

To get the bird's eye view of pollution in each state go back to
map plot, in order to get better picture of pollution in each state,
need to consider the unhealthy days for sensitive groups with AQI 
level 101 to 150.


```{r echo=FALSE, warning=FALSE, Sensitive_day_plot}

states_AQI_map <- ggplot(data = states_AQI, aes(x = long, y = lat, group = group)) +
  geom_polygon() + 
  geom_path(colour = 'grey', linestyle = 2) + 
  coord_map() +
  theme(legend.position = "bottom") +
  geom_text(data = states, aes(x = x, y = y, label = state.abb, group = NULL), 
            size = 4, 
            colour="red") +
  geom_polygon(aes(fill = mean_Sensitive_AQI)) +
  labs(fill = "") +
  guides(fill = guide_colorbar(barwidth = 10, barheight = 0.5)) +
  ggtitle("Percent Sensitive Days of total measured days")

states_AQI_map
```


Explore further based on state area

```{r echo=FALSE, warning=FALSE, Calculate_top_area}
top_area <- states %>% 
  filter(state.area > 100000)

head(top_area[, 1:3])
```

```{r echo=FALSE, warning=FALSE, Plot_based_on_top_area}
top_area_map_in_us <- states_AQI_map + geom_point(data = top_area, 
                                                  inherit.aes = F, aes(x, y), size = 3,
                                                  color = "green", alpha = 0.5) +
  labs(subtitle = "States with area greater than 100000")

top_area_map_in_us
```

The green dot indicates the top biggest states in US


### Reflection

  While starting the project want to plot the pollution levels of various 
states in USA, after exploring the available variables in the datasets the 
direction changed to find the least polluted and most polluted states. 

  The simple univariate challenge gave a lots of thought about
how to handle the data. 

  The next interesting part of finding is the correlation between 
sensitive AQI days with respect to good AQI days and it reflected a strong 
downhill (negative) linear relationship between Good days AQI and Sensitive
days AQI.

  Further exploring the number of days good AQI (value 0 through 50) and 
number of days unhealthy for sensitive AQI (value 51 through 100), 
it was interesting to plot the data in US map, so that we can get the complete
view of pollution in each counties in US. 

 There is lots of room to further explore the same data set for relationship
between ozone and PM2.5 pollutant, identifying which is the most polluting 
component in each state, how the population in each state will impact the 
pollution and vice versa.



